Fractional Types

### 2019-11-22 fracGC/p.tex
trace1
figure for languages and names

@@

%%%%%
\subsection{Choosing Ancillae Values}

@@

In addition to being more expressive, this version has the advantage
of replacing the cumbersome cardinality-based evidence that the type
is not empty by the more intuitive evidence of choosing a value of the
type. The simplications are evident in the Agda formalization:

@@

%%
\subsection{Abstract Machine without Fractionals}

@@

%%
\subsection{Abstract Machine with Fractionals}

@@

%%%%%
\subsection{Pointed and Singleton Types}

Following the development for the toy language above, we create a
lifted version of fractional $\Pi$ with pointed types. In addition to
the explicit pointed types of the form $t$\Acon{\#}$v$, we will need a
multiplicative structure over pointed types, fractional types, and a
special kind of pointed type that includes just one value, a singleton
type. The singleton types will allow the type system to track the flow
of one particular value, which is exactly what is needed to prove the
safety of deallocation:

@@

The combinators in the lifted language consist of all the combinators
in the base language but now operating on pointed types. The mediation
between general pointed types and singleton types is done via
\Acon{return} and \Acon{extract} which as we prove below form a
dual monad/comonad pair. The types for \Acon{$\eta$} and
\Acon{$\epsilon$} are now specialized to guarantee safety of
de-allocation as follows. When applying \Acon{$\eta$} at a pointed type, the
current witness value is put in focus in a singleton type and a gc
process for that particular singleton type is created. To apply this
process using \Acon{$\epsilon$} the very same singleton value must be
the current one.

@@

Exploit dependent types to reify proofs in the type system. Type
checking requires (partial) eval. Once we have a proof can we somehow
go back to \verb|PiFracDyn| and remove the runtime check?

Polymorphic type for ancilla?
\\
First we can inject plain Pi type and value into PiFracDyn:
%% \ednote{Pi/.U is not defined before its use below}
\INJU{}
and the combinators for Pi can also be inject into PiFracDyn in a straightforward way:
\INJcomb{}
The given injection is correct because it does not change the denotation of type
\INJUeq{}
,the value
\INJVeq{}
and the interpretation of PiFracDyn is consistent with Pi:
\INJEvaleq{}
To extract a combinator from PiPointedFrac, first we need to extract the corresponding type for PiFracDyn.
Since every type (except Recip) in PiPointedFrac is pointed type so we can always extract a type for plain
Pi and a value from it.
There is no corresponding type of Recip in plain Pi, but we can map it to the
\AgdaOperator{\AgdaInductiveConstructor{𝟙/}}\AgdaSpace{}\AgdaBound{v} type in PiFracDyn.
Note that for all other type the extracted value is erased from type except for
\AgdaOperator{\AgdaInductiveConstructor{𝟙/}}\AgdaSpace{}\AgdaBound{v}
which we still keep it in type.
\EXTU{}
Once we extract the corresponding type we can extract the combinator, the combinators about monad/comonad
just becomes \AgdaInductiveConstructor{id↔} since the information in type is erased,
for the rest combinators there exist direct corresponding combinators in PiFracDyn:
\EXTUComb{}
Finally we can prove the correctness of the extraction,
the lifted multiplicative combinators have the corresponding combinators in PiFracDyn which represent
the same computation so the proof is just \AgdaInductiveConstructor{refl}.
Because the combinators about monad/comonad does not really
computes so for those case the proof is also \AgdaInductiveConstructor{refl}.
And for \AgdaInductiveConstructor{∙c} we rely on \AgdaFunction{Eval≡}.
The interesting case is \AgdaInductiveConstructor{ε} where the singleton type guarantee that
the dynamic check won't fail.
\EXTeq{}

>>>>

The paper includes a number of languages that express different
points. The three most important languages and their relationships are
highlighted in this diagram:
\begin{center}
\input{lang.tikz}
\end{center}

@@

%%%%%
\subsection{Choosing Ancillae Values: $\Pi/D$}

@@

We will call this language $\Pi/D$ for the fractional extension of
$\Pi$ with a dynamic check. In addition to being more expressive than
the language in the previous subsection, $\Pi/D$ has the advantage of
replacing the cumbersome cardinality-based evidence that the type is
not empty by the more intuitive evidence of choosing a value of the
type. The simplications are evident in the Agda formalization:

@@

%%
\subsection{Abstract Machine for $\Pi$}

@@

%%
\subsection{Abstract Machine for $\Pi/D$}

@@

%%%%%
\subsection{Pointed and Singleton Types: $\Pi/\bullet$}

Following the development for the toy language above, we create a
lifted version of fractional $\Pi$ with pointed types which we call
$\Pi/\bullet$. In addition to the explicit pointed types of the form
$t$\Acon{\#}$v$, we will need a multiplicative structure over pointed
types, fractional types, and a special kind of pointed type that
includes just one value, a singleton type. The singleton types will
allow the type system to track the flow of one particular value, which
is exactly what is needed to prove the safety of deallocation:

@@

The combinators in the lifted language $\Pi/\bullet$ consist of all the combinators
in the core $\Pi$ language but now operating on pointed types. The mediation
between general pointed types and singleton types is done via
\Acon{return} and \Acon{extract} which as we prove below form a
dual monad/comonad pair. The types for \Acon{$\eta$} and
\Acon{$\epsilon$} are now specialized to guarantee safety of
de-allocation as follows. When applying \Acon{$\eta$} at a pointed type, the
current witness value is put in focus in a singleton type and a gc
process for that particular singleton type is created. To apply this
process using \Acon{$\epsilon$} the very same singleton value must be
the current one.

@@

By lifting programs and their evaluation to the type level, we can
naturally leverage the typechecking process to verify properties of
interest, including the safe de-allocation of ancillae. One ``could''
just write programs in that style: indeed the dual nature of proofs
and programs is more and more exploited in languages like the one used
to formalize this paper (Agda). The examples in Sec.~\ref{sec:cat} are
written in that style and they are relatively similar to programs
in the ``normal'' language.

However, it is also often the case than constructive proofs are
further processed to extract a native efficient programs that eschew
all the information needed for the proof invariants. In our case, the
question is whether we can extract a program that runs in the XX but
is guaranteed to be safe and hence can run without the runtime checks
associated with de-allocation sites.

First we can inject plain Pi type and value into PiFracDyn:
%% \ednote{Pi/.U is not defined before its use below}
\INJU{}
and the combinators for Pi can also be inject into PiFracDyn in a straightforward way:
\INJcomb{}
The given injection is correct because it does not change the denotation of type
\INJUeq{}
,the value
\INJVeq{}
and the interpretation of PiFracDyn is consistent with Pi:
\INJEvaleq{}
To extract a combinator from PiPointedFrac, first we need to extract the corresponding type for PiFracDyn.
Since every type (except Recip) in PiPointedFrac is pointed type so we can always extract a type for plain
Pi and a value from it.
There is no corresponding type of Recip in plain Pi, but we can map it to the
\AgdaOperator{\AgdaInductiveConstructor{𝟙/}}\AgdaSpace{}\AgdaBound{v} type in PiFracDyn.
Note that for all other type the extracted value is erased from type except for
\AgdaOperator{\AgdaInductiveConstructor{𝟙/}}\AgdaSpace{}\AgdaBound{v}
which we still keep it in type.
\EXTU{}
Once we extract the corresponding type we can extract the combinator, the combinators about monad/comonad
just becomes \AgdaInductiveConstructor{id↔} since the information in type is erased,
for the rest combinators there exist direct corresponding combinators in PiFracDyn:
\EXTUComb{}
Finally we can prove the correctness of the extraction,
the lifted multiplicative combinators have the corresponding combinators in PiFracDyn which represent
the same computation so the proof is just \AgdaInductiveConstructor{refl}.
Because the combinators about monad/comonad does not really
computes so for those case the proof is also \AgdaInductiveConstructor{refl}.
And for \AgdaInductiveConstructor{∙c} we rely on \AgdaFunction{Eval≡}.
The interesting case is \AgdaInductiveConstructor{ε} where the singleton type guarantee that
the dynamic check won't fail.
\EXTeq{}
